---
title: "class12"
author: "Ella Chang A17902833"
format: pdf
---

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding.

The sample sizes for each genotype were:
A/A: 108 samples
A/G: 233 samples
G/G: 121 samples

The median ORMDL3 expression values for each genotype were:
A/A: 31.25
A/G: 25.06
G/G: 20.07

```{r}
data <- read.table(
  "rs8067378_ENSG00000172057.6.txt",
  header = TRUE,
  stringsAsFactors = FALSE
)


# Sample size per genotype
table(data$geno)

# Median expression per genotype
tapply(data$exp, data$geno, median)

# (Optional) also show summary stats per genotype
by(data$exp, data$geno, summary)
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

From the box plot, it is clear that the expression of ORMDL3 is highest in the A/A genotype, intermediate in the A/G genotype, and lowest in the G/G genotype. This indicates that the G allele is linked with low expression of the ORMDL3 gene. Hence, it can be concluded that the SNP rs8067378 has an effect on the expression of the ORMDL3 gene, and the lowest expression is found in the G/G genotype.
```{r}
boxplot(exp ~ geno,
        data = data,
        xlab = "Genotype",
        ylab = "ORMDL3 Expression",
        main = "ORMDL3 Expression by rs8067378 Genotype")

stripchart(exp ~ geno,
           data = data,
           vertical = TRUE,
           method = "jitter",
           pch = 20,
           col = "gray",
           add = TRUE)
```

